---
title: "VAT RNA-seq: differential expression, pathway enrichment and classification (DG vs NG)"
author: "jmao3cam"
output:
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
---

<!-- ==========================================================================
Notebook : RNA_11_06_run.Rmd
Project  : Multi-omics profiling of visceral adipose tissue (VAT) adipocytes in
           dysglycaemic (DG) vs normoglycaemic (NG) obese women
Layer    : RNA-seq (transcriptomics)
Canonical run : outputs written to 11.06.run/

Inputs (place in the working directory):
  - raw_count.tsv            gene x sample raw counts (ENSEMBL row IDs)
  - human_gene_names.txt     ENSEMBL -> Gene.name annotation
  - new_metatable copy.txt   sequencing metadata (Index, Index.1, New.Grouping, Name)
  - clinical_metadata.xlsx    two-row header clinical table (DG/NG, HbA1c, HOMA-IR, ...)

Key outputs (11.06.run/):
  - DG_vs_NG_all.txt, cpm.txt, DEG up/down tables
  - PCA / tSNE / volcano / snoRNA lollipop PDFs
  - KEGG & GO:BP enrichment tables and dotplots (up/down, ORA on FDR<0.05 set)
  - cv_rf_roc.pdf (cross-validated random-forest ROC)
  - SupplementaryFigure5_snoRNA_QC.pdf

Analysis parameters:
  - DEG (main results) : FDR < 0.05 & |log2FC| > log2(1.5)   -> 210 DEGs (187 up / 23 down)
  - ORA input set      : FDR < 0.05 only (no FC filter)      -> larger set, distinct from above
  - Excluded subjects  : 55, 80, 95, 160, 163, 230, 256, 263, 266, 57, 242
  - MTND1P23 removed from the volcano as a likely technical artefact

NOTE ON REPRODUCIBILITY: run chunks top-to-bottom. `res_df` is created in the
volcano chunk and reused by later chunks (snoRNA diagnostics, RF, DML cross-ref),
so those must be run after it.
=========================================================================== -->

## 1. Data loading + packages

```{r}

# Create the canonical output directory up front so every later ggsave()/write.table()
# has somewhere to write, regardless of which chunk runs first.
dir.create("11.06.run", showWarnings = FALSE, recursive = TRUE)

library(readxl)
library(dplyr)
library(stringr)
library(edgeR)

data  <- read.delim("raw_count.tsv", sep = "\t", row.names = 1)
genes <- read.delim("human_gene_names.txt", sep = "\t", row.names = 1)
meta  <- read.delim("new_metatable copy.txt", sep = "\t")
meta  <- meta[order(meta$Index), ]

colnames(data) <- meta$New.Grouping[match(colnames(data), meta$Index.1)]
data <- data[, !is.na(colnames(data))]

y <- DGEList(counts = data, genes = genes[rownames(data), ])
y$samples$group   <- as.factor(str_sub(meta$Name, 8,  9))
y$samples$subject <- as.factor(str_sub(meta$Name, 11, 20))

remove_subjects <- c(55, 80, 95, 160, 163, 230, 256, 263, 266, 57,242)
keep_idx <- which(!y$samples$subject %in% remove_subjects)
meta <- meta[keep_idx, ]
y    <- y[, keep_idx]
```

## 2. Clinical data loading from excel file

```{r}
library(readxl)
library(dplyr)

# ── Load + clean in one pass ──────────────────────────────────────────────────
clin_raw  <- read_excel("clinical_metadata.xlsx", col_names = FALSE, skip = 0)
top       <- as.character(clin_raw[1, ])
sub       <- as.character(clin_raw[2, ])

for (i in seq_along(top)) {
  if (is.na(top[i]) || top[i] == "NA") top[i] <- top[i - 1]
}

col_names <- ifelse(is.na(sub) | sub == "NA" | sub == top,
                    top, paste(top, sub, sep = "_"))
col_names <- make.names(col_names, unique = TRUE)

clinical  <- read_excel("clinical_metadata.xlsx", skip = 2, col_names = col_names) %>%
  rename(
    sample_id       = Sample.ID,
    group_clin      = Group,
    subject         = Subject,
    sex             = Sex,
    age             = Age,
    height          = Height,
    weight          = Weight,
    bmi             = BMI,
    smoking         = Smoking.Habits,
    cigs_day        = Smoking.Habits_Cigs.day,
    drinking        = Drinking.Habits,
    sbp             = Blood.Pressure_SBP,
    dbp             = Blood.Pressure_DBP,
    heart_rate      = Heart.Rate,
    rx_metformin    = Rx.T2D_Metformin,
    rx_insulin      = Rx.T2D_Insulin,
    rx_dpp4         = Rx.T2D_DPP4.inhib.,
    rx_sglt2        = Rx.T2D_SGLT2.inhib.,
    rx_glp1         = Rx.T2D_GLP.1R.agonists,
    rx_sulfonylurea = Rx.T2D_Sulfonylureas,
    rx_other_t2d    = Rx.T2D_Other.drugs,
    rx_highbp       = Rx.High.BP,
    rx_statins      = Rx.Dyslipidemia,
    rx_other        = Rx.Other,
    hemoglobin      = Hemoglobin,
    glucose         = Glucose,
    hba1c_pct       = HbA1c,
    total_chol      = Total.Cholesterol,
    hdl             = HDL,
    ldl             = LDL,
    triglycerides   = Triglycerides,
    ast             = AST,
    alt             = ALT,
    ggt             = Gama.GT,
    alk_phos        = Alkaline.phosphatase,
    total_protein   = Total.Protein,
    albumin         = Albumine,
    tsh             = TSH,
    insulin_uU      = Insulin,
    peptide_c       = Peptide.C,
    homa_ir         = HOMA.Insulin.Resistance
  ) %>%
  mutate(
    subject    = trimws(as.character(as.integer(subject))),
    group_clin = as.factor(group_clin),
    sex        = as.factor(sex),
    bmi        = as.numeric(bmi),
    age        = as.numeric(age),
    glucose    = as.numeric(glucose),
    hba1c_pct  = as.numeric(hba1c_pct),
    homa_ir    = as.numeric(homa_ir),
    sbp        = as.numeric(sbp),
    dbp        = as.numeric(dbp),
    treatment  = as.factor(ifelse(
      rx_metformin == 1 | rx_insulin == 1 | rx_dpp4 == 1 | rx_sglt2 == 1 |
      rx_glp1 == 1 | rx_sulfonylurea == 1 | rx_other_t2d == 1,
      "Treatment", "No treatment"
    ))
  )

# ── Verify ────────────────────────────────────────────────────────────────────
cat("Subjects:", head(clinical$subject), "\n")
cat("Treatment breakdown:\n")
print(table(clinical$treatment, clinical$group_clin))
```

### Treatment vs no treatment group separation

```{r}
clinical <- clinical %>%
  mutate(
    treatment = as.factor(ifelse(
      rx_metformin == 1 | rx_insulin == 1 | rx_dpp4 == 1 | rx_sglt2 == 1 |
      rx_glp1 == 1 | rx_sulfonylurea == 1 | rx_other_t2d == 1 ,
      "Treatment", "No treatment"
    ))
  )

cat("Treatment group breakdown:\n")
print(table(clinical$treatment, clinical$group_clin))
```

## 3. Filter and normalisation of the data

gene must be expressed in at least as many samples as the smallest group (32)

```{r}
# 7. Filter, normalise, dispersion 
keep <- rowSums(cpm(y) >= 1) >= 32
y    <- y[keep, , keep.lib.size = FALSE]
y    <- calcNormFactors(y)
nrow(y)

design <- model.matrix(~0 + y$samples$group)
colnames(design) <- c("DG", "NG")

```

```{r}
## PCA on filtered, normalised RNA-seq data

library(ggplot2)

logcpm_pca <- cpm(y, log = TRUE, prior.count = 1)
mat_pca    <- t(logcpm_pca)

pca_out <- prcomp(mat_pca, center = TRUE, scale. = FALSE)
var_explained <- round(100 * (pca_out$sdev^2) / sum(pca_out$sdev^2), 1)

plot_df_pca <- data.frame(
  PC1     = pca_out$x[, 1],
  PC2     = pca_out$x[, 2],
  group   = y$samples$group,
  subject = y$samples$subject
)

if (exists("clinical")) {
  plot_df_pca$subject <- as.character(plot_df_pca$subject)
  clinical$subject     <- as.character(clinical$subject)
  plot_df_pca <- left_join(plot_df_pca, clinical[, c("subject", "treatment")], by = "subject")
  plot_df_pca$treatment[is.na(plot_df_pca$treatment)] <- "Unknown"
}

x_lim <- max(abs(plot_df_pca$PC1), na.rm = TRUE) * 1.05
y_lim <- max(abs(plot_df_pca$PC2), na.rm = TRUE) * 1.05

p_pca <- ggplot(plot_df_pca, aes(x = PC1, y = PC2, colour = group)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0"),
                      labels = c("DG" = "Dysglycaemic", "NG" = "Normoglycaemic")) +
  scale_x_continuous(limits = c(-x_lim, x_lim)) +
  scale_y_continuous(limits = c(-y_lim, y_lim)) +
  labs(title  = "PCA - VAT RNA-seq (filtered, TMM-normalised)",
       x      = paste0("PC1 (", var_explained[1], "%)"),
       y      = paste0("PC2 (", var_explained[2], "%)"),
       colour = "Group") +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    plot.subtitle   = element_text(hjust = 0.5, colour = "grey40", size = 10),
    axis.title      = element_text(size = 12),
    axis.text       = element_text(size = 10.5, colour = "grey20"),
    legend.text     = element_text(size = 10.5),
    plot.margin     = ggplot2::margin(10, 20, 10, 10)
  )

print(p_pca)
ggsave("11.06.run/pca_DG_vs_NG.pdf", p_pca, width = 10, height = 6)
```


## tSNE on filtered, normalised RNA-seq data
Shows group separation
```{r}
## tSNE on filtered, normalised RNA-seq data

library(Rtsne)
library(ggplot2)

# ── Get log2-CPM matrix (samples as rows) ───────────────────────────────────
logcpm <- cpm(y, log = TRUE, prior.count = 1)
mat    <- t(logcpm)   # samples x genes

# ── Run tSNE ────────────────────────────────────────────────────────────────
set.seed(20)
tsne_out <- Rtsne(mat,
                  dims        = 2,
                  perplexity  = 10,   # ~sqrt(n_samples); adjust if needed
                  max_iter    = 1000,
                  pca         = TRUE,
                  pca_center  = TRUE,
                  pca_scale   = FALSE,
                  verbose     = FALSE)

# ── Build plot data frame ────────────────────────────────────────────────────
plot_df <- data.frame(
  tSNE1   = tsne_out$Y[, 1],
  tSNE2   = tsne_out$Y[, 2],
  group   = y$samples$group,
  subject = y$samples$subject
)

# Optionally merge treatment info if clinical is available
if (exists("clinical")) {
  plot_df$subject <- as.character(plot_df$subject)
  clinical$subject <- as.character(clinical$subject)
  plot_df <- left_join(plot_df, clinical[, c("subject", "treatment")], by = "subject")
  plot_df$treatment[is.na(plot_df$treatment)] <- "Unknown"
}

# ── Plot ─────────────────────────────────────────────────────────────────────
p <- ggplot(plot_df, aes(x = tSNE1, y = tSNE2, colour = group)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0"),
                      labels = c("DG" = "Dysglycaemic", "NG" = "Normoglycaemic")) +
  labs(title   = "tSNE - VAT RNA-seq (filtered, TMM-normalised)",
       x       = "tSNE 1",
       y       = "tSNE 2",
       colour  = "Group") +
  theme_bw(base_size = 13) +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))

print(p)

# ── Optional: colour by treatment status ────────────────────────────────────
if ("treatment" %in% colnames(plot_df)) {
  p2 <- ggplot(plot_df, aes(x = tSNE1, y = tSNE2,
                             colour = group, shape = treatment)) +
    geom_point(size = 3, alpha = 0.85) +
    scale_colour_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0")) +
    labs(title  = "tSNE - coloured by group, shaped by treatment",
         colour = "Group", shape = "Treatment") +
    theme_bw(base_size = 13) +
    theme(legend.position = "right",
          plot.title = element_text(hjust = 0.5))
  print(p2)
}
```

## Statistical tests 
### Uses glmFit/glmLRT

```{r}
# ── 8. Fit & test DG vs NG ────────────────────────────────────────────────────
y    <- estimateDisp(y, design)
fit  <- glmFit(y, design)
DG_vs_NG <- glmLRT(fit, contrast = c(1, -1))

cat("\nDEG table:\n")
print(table(decideTests(DG_vs_NG)))

write.table(topTags(DG_vs_NG, n = Inf),
            "DG_vs_NG.txt", sep = "\t", quote = FALSE, col.names = NA)

# ── 9. CPM export ─────────────────────────────────────────────────────────────
log_cpm <- cpm(y, log = FALSE)
log_cpm <- log2(log_cpm + 1)
# colnames are already correct from y — don't overwrite with meta
log_cpm <- merge(log_cpm, genes, by = 0, all.x = TRUE)
write.table(log_cpm, "cpm.txt", sep = "\t", quote = FALSE, col.names = NA)
#save the table with all the data 
res <- topTags(DG_vs_NG, n = Inf)$table
write.table(res, "DG_vs_NG_all.txt", sep = "\t", quote = FALSE, col.names = NA)
```
### All DEGs

```{r}
all_degs <- res %>%
  filter(FDR < 0.05 & abs(logFC) > log2(1.5)) %>%
  arrange(FDR) %>%
  mutate(
    FC        = round(logFC, 2),
    FDR       = formatC(FDR, format = "e", digits = 2),
    Direction = ifelse(logFC > 0, "Up", "Down")
  ) %>%
  dplyr::select(Gene = Gene.name, FC, Direction, FDR)

print(all_degs, row.names = FALSE)
cat("Total DEGs:", nrow(all_degs), "\n")
```

### Gene Upregulation or Downregulation

```{r}
# ── Create output folder ──────────────────────────────────────────────────────
dir.create("11.06.run", showWarnings = FALSE)

# ── DEG classification ────────────────────────────────────────────────────────
res <- topTags(DG_vs_NG, n = Inf)$table

res$DEG <- ifelse(res$FDR < 0.05 & abs(res$logFC) > 0.58,
                  ifelse(res$logFC > 0, "Upregulated", "Downregulated"),
                  "Not Significant")

# ── Summary table ─────────────────────────────────────────────────────────────
print(table(res$DEG))

# ── Split up/down ─────────────────────────────────────────────────────────────
up   <- res[res$DEG == "Upregulated",   ] |> arrange(FDR)
down <- res[res$DEG == "Downregulated", ] |> arrange(FDR)

cat("Upregulated:", nrow(up), "| Downregulated:", nrow(down), "\n")

# ── Export ────────────────────────────────────────────────────────────────────
write.table(up,                                  "11.06.run/DG_vs_NG_upregulated.txt",   sep = "\t", quote = FALSE, col.names = NA)
write.table(down,                                "11.06.run/DG_vs_NG_downregulated.txt", sep = "\t", quote = FALSE, col.names = NA)
write.table(res[res$DEG != "Not Significant", ], "11.06.run/DG_vs_NG_DEGs_all.txt",      sep = "\t", quote = FALSE, col.names = NA)
```

## KEGG and GO pathway enrichment analysis

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# ── 1. Get significant DEGs ───────────────────────────────────────────────────
sig_genes <- res[res$FDR < 0.05, ]
sig_genes$ENSEMBL <- rownames(sig_genes)

# ── 2. Map ENSEMBL to Entrez IDs ─────────────────────────────────────────────
entrez <- bitr(sig_genes$ENSEMBL,
               fromType = "ENSEMBL",
               toType   = "ENTREZID",
               OrgDb    = org.Hs.eg.db)

sig_merged <- merge(sig_genes, entrez, by = "ENSEMBL")
cat("DEGs mapped to Entrez IDs:", nrow(sig_merged), "\n")

# ── 3. KEGG enrichment ────────────────────────────────────────────────────────
kegg_res <- enrichKEGG(
  gene         = sig_merged$ENTREZID,
  organism     = "hsa",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)

cat("Significant KEGG pathways:", nrow(kegg_res@result[kegg_res@result$p.adjust < 0.05, ]), "\n")
write.table(as.data.frame(kegg_res), "11.06.run/KEGG_enrichment_results.txt", sep = "\t", quote = FALSE, row.names = FALSE)

# ── 4. GO Biological Process enrichment ──────────────────────────────────────
go_bp <- enrichGO(
  gene         = sig_merged$ENTREZID,
  OrgDb        = org.Hs.eg.db,
  ont          = "BP",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable     = TRUE
)

write.table(as.data.frame(go_bp), "11.06.run/GO_BP_enrichment_results.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```

### KEGG and GO tables

```{r}
library(knitr)
library(kableExtra)

# ── KEGG table ────────────────────────────────────────────────────────────────
kegg_top20 <- as.data.frame(kegg_res) %>%
  arrange(p.adjust) %>%
  head(20) %>%
  mutate(
    Rank      = row_number(),
    FDR       = formatC(p.adjust, format = "e", digits = 2),
    GeneRatio = GeneRatio
  ) %>%
  dplyr::select(Rank, Pathway = Description, GeneRatio, FDR)

# ── GO BP table ───────────────────────────────────────────────────────────────
go_top20 <- as.data.frame(go_bp) %>%
  arrange(p.adjust) %>%
  head(20) %>%
  mutate(
    Rank      = row_number(),
    FDR       = formatC(p.adjust, format = "e", digits = 2),
    GeneRatio = GeneRatio
  ) %>%
  dplyr::select(Rank, Pathway = Description, GeneRatio, FDR)

# ── Print as styled tables ────────────────────────────────────────────────────
kegg_top20 %>%
  kbl(caption = "Top 20 KEGG Pathways — DG vs NG", align = c("c","l","c","c")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE,
    font_size         = 12
  ) %>%
  row_spec(0, bold = TRUE, background = "#2C3E50", color = "white") %>%
  column_spec(1, width = "2em", bold = TRUE) %>%
  column_spec(2, width = "20em") %>%
  save_kable("11.06.run/KEGG_top20_table.html")

go_top20 %>%
  kbl(caption = "Top 20 GO Biological Processes — DG vs NG", align = c("c","l","c","c")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE,
    font_size         = 12
  ) %>%
  row_spec(0, bold = TRUE, background = "#2C3E50", color = "white") %>%
  column_spec(1, width = "2em", bold = TRUE) %>%
  column_spec(2, width = "20em") %>%
  save_kable("11.06.run/GO_BP_top20_table.html")
```

## Volcano plot of DEGs 

```{r}
library(ggrepel)
res_df <- as.data.frame(topTags(DG_vs_NG, n = Inf)$table)
res_df[] <- lapply(res_df, function(x) if (is(x, "Rle")) as.vector(x) else x)
res_df <- res_df[res_df$Gene.name != "MTND1P23", ]

volcano_df <- res_df %>%
  mutate(
    neg_log10_FDR = -log10(FDR),
    significance  = case_when(
      FDR < 0.05 & logFC >  log2(1.5) ~ "Up",
      FDR < 0.05 & logFC < -log2(1.5) ~ "Down",
      TRUE                             ~ "NS"
    )
  )

top_fc_genes <- volcano_df %>%
  filter(significance != "NS") %>%
  group_by(significance) %>%
  arrange(desc(abs(logFC))) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  pull(Gene.name)

volcano_df <- volcano_df %>%
  mutate(label = ifelse(Gene.name %in% top_fc_genes, Gene.name, NA_character_))

n_up   <- sum(volcano_df$significance == "Up")
n_down <- sum(volcano_df$significance == "Down")

# symmetric x-limits with a bit of breathing room
x_lim <- max(abs(volcano_df$logFC), na.rm = TRUE) * 1.05

p_volcano <- ggplot(volcano_df, aes(x = logFC, y = neg_log10_FDR, colour = significance)) +
  geom_point(aes(size = significance), alpha = 0.6) +
  geom_text_repel(
    aes(label = label),
    size          = 3,
    max.overlaps  = 20,
    box.padding   = 0.4,
    segment.color = "grey50",
    na.rm         = TRUE
  ) +
  geom_vline(xintercept = c(-log2(1.5), log2(1.5)),
             linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  scale_colour_manual(
    values = c("Up" = "#D94F3D", "Down" = "#4C72B0", "NS" = "grey70"),
    labels = c("Up"   = paste0("Up (", n_up, ")"),
               "Down" = paste0("Down (", n_down, ")"),
               "NS"   = "Not significant")
  ) +
  scale_size_manual(values = c("Up" = 1.8, "Down" = 1.8, "NS" = 0.8),
                    guide = "none") +
  scale_x_continuous(limits = c(-x_lim, x_lim)) +
  coord_cartesian(clip = "off") +
  labs(
    title    = "Differential gene expression - DG vs NG",
    subtitle = "FDR < 0.05 & |FC| > 1.5",
    x        = "log2 Fold Change (DG/NG)",
    y        = "-log10(FDR)",
    colour   = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position   = c(0.08, 0.88),
    legend.background = element_blank(),
    plot.title        = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle     = element_text(hjust = 0.5, colour = "grey40", size = 13),
    axis.title        = element_text(size = 15),
    axis.text         = element_text(size = 12.5, colour = "grey20"),
    legend.text       = element_text(size = 12.5),
    plot.margin       = margin(10, 20, 10, 10)
  )

print(p_volcano)

ggsave("11.06.run/volcano_DG_vs_NG.pdf", p_volcano, width = 12, height = 6)
```


```{r}
library(dplyr)
library(gridExtra)
library(grid)

kegg_selected <- c("Carbon metabolism", "Citrate cycle (TCA cycle)",
                   "AMPK signaling pathway", "Diabetic cardiomyopathy", "Insulin signaling pathway")

go_selected <- c("generation of precursor metabolites and energy",
                 "cellular respiration",
                 "MHC class II protein complex assembly")

kegg_table <- as.data.frame(kegg_res) %>%
  filter(Description %in% kegg_selected) %>%
  arrange(p.adjust) %>%
  mutate(FDR = formatC(p.adjust, format = "e", digits = 2),
         Database = "KEGG") %>%
  dplyr::select(Database, Pathway = Description, GeneRatio, FDR)

go_table <- as.data.frame(go_bp) %>%
  filter(Description %in% go_selected) %>%
  arrange(p.adjust) %>%
  mutate(FDR = formatC(p.adjust, format = "e", digits = 2),
         Database = "GO:BP") %>%
  dplyr::select(Database, Pathway = Description, GeneRatio, FDR)

combined <- bind_rows(kegg_table, go_table)

pdf("11.06.run/selected_pathways_table.pdf", width = 11, height = 5)

tt <- ttheme_default(
  core = list(
    bg_params = list(
      fill = c(rep(c("#EAF2FB", "white"), length.out = nrow(kegg_table)),
               rep(c("#EAF8F1", "white"), length.out = nrow(go_table))),
      col = "grey85"
    ),
    fg_params = list(fontsize = 10, hjust = 0, x = 0.05)
  ),
  colhead = list(
    bg_params = list(fill = "#2C3E50", col = NA),
    fg_params = list(fontsize = 11, fontface = "bold", col = "white",
                     hjust = 0, x = 0.05)
  )
)

tbl <- tableGrob(combined, rows = NULL, theme = tt)

# Add section separating line between KEGG and GO
grid.newpage()
grid.arrange(
  tbl,
  top = textGrob("Enriched Pathways - DG vs NG",
                 gp = gpar(fontsize = 14, fontface = "bold")),
  padding = unit(1, "cm")
)

dev.off()
```
```{r}
top_gene   <- "SNORA26"
gene_ensid <- rownames(res_df)[res_df$Gene.name == top_gene][1]

logcpm <- cpm(y, log = TRUE, prior.count = 1)

plot_df_deg <- data.frame(
  logCPM = as.numeric(logcpm[gene_ensid, ]),
  Group  = y$samples$group
)

ggplot(plot_df_deg, aes(x = Group, y = logCPM, fill = Group)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.7,
               colour = "grey30", linewidth = 0.6) +
  geom_jitter(aes(colour = Group), width = 0.12, size = 2.5,
              alpha = 0.85, show.legend = FALSE) +
  scale_fill_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0"),
                    labels = c("DG" = "Dysglycaemic", "NG" = "Normoglycaemic")) +
  scale_colour_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0")) +
  scale_x_discrete(labels = c("DG" = "Dysglycaemic", "NG" = "Normoglycaemic")) +
  labs(
    title = "SNORA26 Expression",
    x     = NULL,
    y     = "log2-CPM",
    fill  = NULL
  ) +
  theme_classic(base_size = 16) +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold.italic", size = 17),
    legend.position = "none",
    axis.text.x     = element_text(size = 14, colour = "grey20"),
    axis.text.y     = element_text(size = 13),
    axis.title.y    = element_text(size = 14)
  )

ggsave("11.06.run/boxplot_SNORA26_expression.pdf", width = 4.5, height = 5.5)
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# ── Use corrected threshold ────────────────────────────────────────────────
sig_up   <- res_df[res_df$FDR < 0.05 & res_df$logFC > 0, ]
sig_down <- res_df[res_df$FDR < 0.05 & res_df$logFC < 0, ]

cat("Upregulated DEGs:  ", nrow(sig_up), "\n")
cat("Downregulated DEGs:", nrow(sig_down), "\n")

# ── Map ENSEMBL to Entrez ─────────────────────────────────────────────────
map_entrez <- function(df) {
  df$ENSEMBL <- rownames(df)
  entrez <- bitr(df$ENSEMBL, fromType = "ENSEMBL",
                 toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  merge(df, entrez, by = "ENSEMBL")
}

up_mapped   <- map_entrez(sig_up)
down_mapped <- map_entrez(sig_down)

# ── KEGG enrichment ───────────────────────────────────────────────────────
kegg_up <- enrichKEGG(gene = up_mapped$ENTREZID, organism = "hsa",
                      pvalueCutoff = 0.05, qvalueCutoff = 0.2)

kegg_down <- enrichKEGG(gene = down_mapped$ENTREZID, organism = "hsa",
                        pvalueCutoff = 0.05, qvalueCutoff = 0.2)

# ── GO:BP enrichment ──────────────────────────────────────────────────────
go_up <- enrichGO(gene = up_mapped$ENTREZID, OrgDb = org.Hs.eg.db,
                  ont = "BP", pvalueCutoff = 0.05, qvalueCutoff = 0.2,
                  readable = TRUE)

go_down <- enrichGO(gene = down_mapped$ENTREZID, OrgDb = org.Hs.eg.db,
                    ont = "BP", pvalueCutoff = 0.05, qvalueCutoff = 0.2,
                    readable = TRUE)

# ── Print summaries ───────────────────────────────────────────────────────
cat("\nKEGG - Upregulated:", nrow(as.data.frame(kegg_up)), "pathways\n")
cat("KEGG - Downregulated:", nrow(as.data.frame(kegg_down)), "pathways\n")
cat("GO:BP - Upregulated:", nrow(as.data.frame(go_up)), "terms\n")
cat("GO:BP - Downregulated:", nrow(as.data.frame(go_down)), "terms\n")

# ── Top 10 per direction ──────────────────────────────────────────────────
cat("\nTop KEGG - Upregulated:\n")
print(as.data.frame(kegg_up) %>% arrange(p.adjust) %>% head(10) %>%
        dplyr::select(Description, GeneRatio, p.adjust))

cat("\nTop KEGG - Downregulated:\n")
print(as.data.frame(kegg_down) %>% arrange(p.adjust) %>% head(10) %>%
        dplyr::select(Description, GeneRatio, p.adjust))

cat("\nTop GO:BP - Upregulated:\n")
print(as.data.frame(go_up) %>% arrange(p.adjust) %>% head(10) %>%
        dplyr::select(Description, GeneRatio, p.adjust))

cat("\nTop GO:BP - Downregulated:\n")
print(as.data.frame(go_down) %>% arrange(p.adjust) %>% head(10) %>%
        dplyr::select(Description, GeneRatio, p.adjust))

# ── Export ────────────────────────────────────────────────────────────────
write.table(as.data.frame(kegg_up),   "11.06.run/KEGG_upregulated.txt",   sep = "\t", quote = FALSE, row.names = FALSE)
write.table(as.data.frame(kegg_down), "11.06.run/KEGG_downregulated.txt", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(as.data.frame(go_up),     "11.06.run/GO_BP_upregulated.txt",  sep = "\t", quote = FALSE, row.names = FALSE)
write.table(as.data.frame(go_down),   "11.06.run/GO_BP_downregulated.txt",sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r}
library(dplyr)
library(gridExtra)
library(grid)

# ── Selected pathways from corrected analysis ──────────────────────────────
kegg_selected_up <- c("Systemic lupus erythematosus", "Asthma",
                      "Intestinal immune network for IgA production",
                      "Antigen processing and presentation",
                      "Type I diabetes mellitus")

go_selected_up <- c("antigen processing and presentation of exogenous peptide antigen",
                    "antigen processing and presentation of peptide antigen via MHC class II",
                    "MHC class II protein complex assembly")

kegg_table <- as.data.frame(kegg_up) %>%
  filter(Description %in% kegg_selected_up) %>%
  arrange(p.adjust) %>%
  mutate(FDR      = formatC(p.adjust, format = "e", digits = 2),
         Database = "KEGG") %>%
  dplyr::select(Database, Pathway = Description, GeneRatio, FDR)

go_table <- as.data.frame(go_up) %>%
  filter(Description %in% go_selected_up) %>%
  arrange(p.adjust) %>%
  mutate(FDR      = formatC(p.adjust, format = "e", digits = 2),
         Database = "GO:BP") %>%
  dplyr::select(Database, Pathway = Description, GeneRatio, FDR)

combined <- bind_rows(kegg_table, go_table)

# ── Table theme ────────────────────────────────────────────────────────────
tt <- ttheme_default(
  core = list(
    bg_params = list(
      fill = c(rep(c("#EAF2FB", "white"), length.out = nrow(kegg_table)),
               rep(c("#EAF8F1", "white"), length.out = nrow(go_table))),
      col = "grey85"
    ),
    fg_params = list(fontsize = 10, hjust = 0, x = 0.05)
  ),
  colhead = list(
    bg_params = list(fill = "#2C3E50", col = NA),
    fg_params = list(fontsize = 11, fontface = "bold", col = "white",
                     hjust = 0, x = 0.05)
  )
)

pdf("11.06.run/selected_pathways_table_corrected.pdf", width = 11, height = 5)
grid.newpage()
grid.arrange(
  tableGrob(combined, rows = NULL, theme = tt),
  top = textGrob("Enriched Pathways - Upregulated DEGs (FDR<0.05)",
                 gp = gpar(fontsize = 14, fontface = "bold")),
  padding = unit(1, "cm")
)
dev.off()
```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)

make_enrich_plot <- function(enrich_obj, title, fill_colour, n = 10) {
  df <- as.data.frame(enrich_obj) %>%
    arrange(p.adjust) %>%
    head(n) %>%
    mutate(
      Description   = stringr::str_wrap(Description, width = 40),
      Description   = factor(Description, levels = rev(Description)),
      gene_count    = as.numeric(sub("/.*", "", GeneRatio)),
      neg_log10_fdr = -log10(p.adjust)
    )
  
  ggplot(df, aes(x = neg_log10_fdr, y = Description)) +
    geom_col(fill = fill_colour, alpha = 0.85, width = 0.7) +
    geom_point(aes(size = gene_count), colour = "grey20", alpha = 0.8) +
    scale_size_continuous(name = "Gene count", range = c(2, 6)) +
    labs(title = title, x = "-log10(FDR)", y = NULL) +
    theme_classic(base_size = 11) +
    theme(
      plot.title   = element_text(face = "bold", size = 11, hjust = 0),
      axis.text.y  = element_text(size = 9),
      axis.text.x  = element_text(size = 9),
      legend.position = "right"
    )
}

p1 <- make_enrich_plot(kegg_up,   "KEGG — Upregulated",   "#E8534A")
p2 <- make_enrich_plot(kegg_down, "KEGG — Downregulated", "#4C72B0")
p3 <- make_enrich_plot(go_up,     "GO:BP — Upregulated",  "#E8534A")
p4 <- make_enrich_plot(go_down,   "GO:BP — Downregulated","#4C72B0")

combined <- (p1 | p2) / (p3 | p4) +
  plot_annotation(
    title    = "Pathway enrichment — DG vs NG",
    subtitle = "Top 10 terms by FDR, ORA (FDR < 0.05)",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 13, hjust = 0.5),
      plot.subtitle = element_text(size = 10, hjust = 0.5, colour = "grey40")
    )
  )

ggsave("11.06.run/enrichment_plots.pdf", combined, width = 14, height = 12)
```


```{r}
library(ggplot2)
library(dplyr)

# ── Extract all snoRNAs from DEGs ─────────────────────────────────────────
snorna_df <- res_df %>%
  filter(FDR < 0.05 & logFC > log2(1.5)) %>%
  filter(grepl("^SNORA|^SNORD|^SCARNA", Gene.name)) %>%
  arrange(desc(logFC)) %>%
  mutate(Gene.name = factor(Gene.name, levels = Gene.name[order(logFC)]))

cat("Total snoRNAs in upregulated DEGs:", nrow(snorna_df), "\n")
print(snorna_df[, c("Gene.name", "logFC", "FDR")])

# ── Lollipop plot ─────────────────────────────────────────────────────────
ggplot(snorna_df, aes(x = logFC, y = Gene.name)) +
  geom_segment(aes(x = 0, xend = logFC, yend = Gene.name),
               linewidth = 0.8, colour = "#E8534A") +
  geom_point(size = 3.5, colour = "#E8534A") +
  geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey30") +
  scale_x_continuous(limits = c(0, NA),
                     breaks = seq(0, 2, by = 0.5)) +
  labs(
    title    = "Upregulated snoRNAs in DG VAT",
    subtitle = "FDR < 0.05 & |FC| > 1.5",
    x        = "log2 Fold Change (DG/NG)",
    y        = NULL
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title    = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey40", size = 10),
    axis.text.y   = element_text(face = "italic", size = 11),
    axis.text.x   = element_text(size = 11),
    axis.title.x  = element_text(size = 12),
    panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3)
  )

ggsave("11.06.run/lollipop_snoRNAs_upregulated.pdf", width = 6, height = 7)
```

```{r cv-rf-rnaseq, warning=FALSE, message=FALSE}
library(randomForest)
library(pROC)

set.seed(42)

n_folds   <- 5
n_repeats <- 10
n_top_genes <- 50   # same number as your original top-50-DEG classifier

group_labels <- y$samples$group  # factor: DG / NG
n_samples    <- length(group_labels)

# Storage for pooled out-of-fold predictions across all repeats
all_true  <- character(0)
all_probs <- numeric(0)

for (rep_i in seq_len(n_repeats)) {

  set.seed(100 + rep_i)

  # Stratified fold assignment: shuffle within each group separately so
  # DG/NG proportions stay balanced across folds despite the 33/32 split
  fold_id <- integer(n_samples)
  for (g in levels(group_labels)) {
    idx <- which(group_labels == g)
    idx <- sample(idx)  # shuffle
    fold_id[idx] <- rep(seq_len(n_folds), length.out = length(idx))
  }

  for (k in seq_len(n_folds)) {
    test_idx  <- which(fold_id == k)
    train_idx <- which(fold_id != k)

    # ── Re-run DE testing using ONLY the training fold ─────────────────────
    y_train <- y[, train_idx]
    y_train <- calcNormFactors(y_train)  # re-normalise on training data only

    design_train <- model.matrix(~0 + y_train$samples$group)
    colnames(design_train) <- levels(y_train$samples$group)

    y_train <- estimateDisp(y_train, design_train)
    fit_train <- glmFit(y_train, design_train)
    lrt_train <- glmLRT(fit_train, contrast = c(1, -1))

    top_genes <- rownames(topTags(lrt_train, n = n_top_genes)$table)

    # ── Build train/test feature matrices from the SAME normalised CPMs ────
    logcpm_all <- cpm(y, log = TRUE, prior.count = 1)

    train_x <- t(logcpm_all[top_genes, train_idx, drop = FALSE])
    test_x  <- t(logcpm_all[top_genes, test_idx,  drop = FALSE])

    train_y <- factor(group_labels[train_idx])  # force factor, in case it was coerced
    test_y  <- factor(group_labels[test_idx])

    rf_fold <- randomForest::randomForest(x = train_x, y = train_y, ntree = 500)

    # Diagnostic (uncomment if the error recurs, to confirm root cause):
    # cat("class:", class(rf_fold), "| type:", rf_fold$type, "\n")

    # Call the classification predict method explicitly rather than the
    # bare generic predict() — this avoids dispatch ambiguity if another
    # loaded package (clusterProfiler/AnnotationDbi both loaded earlier in
    # this notebook mask a number of generics) has shadowed the intended
    # predict.randomForest method.
    fold_probs <- randomForest:::predict.randomForest(rf_fold, newdata = test_x,
                                                       type = "prob")[, "DG"]

    all_true  <- c(all_true,  as.character(test_y))
    all_probs <- c(all_probs, fold_probs)
  }
}

# ── Pooled ROC/AUC across every held-out prediction from every fold/repeat ──
roc_cv <- pROC::roc(all_true, all_probs, levels = c("NG", "DG"), direction = "<")

cat("Cross-validated RNA-seq RF AUC (top", n_top_genes, "genes, refit per fold):",
    round(auc(roc_cv), 3), "\n")
cat("(", n_repeats, "repeats x", n_folds, "folds =", length(all_true), "pooled predictions)\n")

pdf(file.path("11.06.run", "cv_rf_roc.pdf"), width = 5, height = 5)
plot(roc_cv, col = "#D94F3D", lwd = 2,
     main = paste0("RNA-seq CV RF (AUC = ", round(auc(roc_cv), 3), ")"))
dev.off()

cat("\nComparison:\n")
cat("  Top-50-DEG RF, NOT cross-validated (leakage-prone): 0.873\n")
cat("  Top-50-DEG RF, cross-validated (fair):              ", round(auc(roc_cv), 3), "\n")
cat("  MOFA all 15 factors:                                 0.750\n")
cat("  MOFA Factors 1+2+3:                                  0.769\n")
```


```{r}
library(dplyr)

n_deg_total     <- sum(res_df$is_DEG_up, na.rm = TRUE)
n_snorna_tested <- sum(res_df$is_snoRNA, na.rm = TRUE)

# ── 1. Define snoRNA gene sets ──────────────────────────────────────────────
# Background = all genes tested in the DE model (res_df), not the whole genome,
# since untested genes could never have been called DEGs in the first place
res_df <- res_df %>%
  mutate(is_snoRNA = grepl("^SNORA|^SNORD|^SCARNA", Gene.name))

res_df <- res_df %>%
  mutate(is_DEG_up = FDR < 0.05 & logFC > log2(1.5))

# ── 2. Build 2x2 contingency table ──────────────────────────────────────────
# Rows: snoRNA vs non-snoRNA | Columns: DEG (up) vs not DEG
contingency <- table(
  snoRNA = res_df$is_snoRNA,
  DEG    = res_df$is_DEG_up
)

print(contingency)

# ── 3. Chi-square test ───────────────────────────────────────────────────────
chisq_result <- chisq.test(contingency)
print(chisq_result)

# ── 4. Check expected counts (chi-square assumption: all expected >= 5) ────
cat("\nExpected counts:\n")
print(chisq_result$expected)

# If any expected count < 5, Fisher's exact test is more appropriate
if (any(chisq_result$expected < 5)) {
  cat("\nWarning: some expected counts < 5 — chi-square approximation may be unreliable.\n")
  cat("Falling back to Fisher's exact test:\n")
  fisher_result <- fisher.test(contingency)
  print(fisher_result)
}

# ── 5. Summary stats ─────────────────────────────────────────────────────────
n_snorna_deg    <- sum(res_df$is_snoRNA & res_df$is_DEG_up)
n_deg_total     <- sum(res_df$is_DEG_up)
n_snorna_tested <- sum(res_df$is_snoRNA)
n_tested_total  <- nrow(res_df)

pct_deg_snorna   <- 100 * n_snorna_deg / n_deg_total
pct_bg_snorna    <- 100 * n_snorna_tested / n_tested_total

cat("\n── Summary ──\n")
cat("snoRNAs among upregulated DEGs: ", n_snorna_deg, "/", n_deg_total,
    " (", round(pct_deg_snorna, 1), "%)\n", sep = "")
cat("snoRNAs among all tested genes: ", n_snorna_tested, "/", n_tested_total,
    " (", round(pct_bg_snorna, 1), "%)\n", sep = "")
cat("Fold enrichment: ", round(pct_deg_snorna / pct_bg_snorna, 2), "x\n", sep = "")

```

```{r}
library(dplyr)

# ── 1. TSS-filtered candidates (read off Figure panels A & B; only PELP1
# and ZNF423 delta-beta values are confirmed from earlier in this thread —
# fill in the rest once you can access RRBS_locus_DG_vs_NG.txt /
# RRBS_gene_level_DG_vs_NG.txt directly) ────────────────────────────────────
tss_filtered <- data.frame(
  gene       = c("LINC00682", "CSTF2", "PELP1", "H3C1", "ZNF423",
                 "SMU1", "UBE2E4P", "IRX4", "LINC01354"),
  delta_beta = c(NA, NA, 0.088, NA, 0.10,
                 NA, NA, NA, NA),
  PValue     = NA_real_,
  FDR        = NA_real_,
  direction  = c("Hypo", "Hypo", "Hyper", "Hyper", "Hyper",
                 "Hyper", "Hyper", "Hyper", "Hyper"),
  source     = c("TSS locus-level (exploratory)", "TSS locus-level (exploratory)",
                 "TSS locus-level (exploratory)", "TSS locus-level (exploratory)",
                 "TSS locus-level + gene-level (primary)", "TSS locus-level (exploratory)",
                 "TSS locus-level (exploratory)", "TSS gene-level (exploratory)",
                 "TSS gene-level (exploratory)")
)

# ── 2. Whole-genome exploratory candidates — full list, hardcoded directly
# from dml_exploratory_wholegenome.txt (48 loci: 43 hyper, 5 hypo) ──────────
wg_filtered <- data.frame(
  gene = c("S100A10","CROCCP3","FGFR2","PDLIM1","CRTAC1","LOC105369509",
           "OAF","OAF","OAF","OAF","GRAMD1B","LINC01234","PPM1H","ATP11A",
           "MBNL2","MBNL2","CDH5","LOC105371485","PELP1","CYTH1","LINC01970",
           "ANO8","LAPTM4A","LAPTM4A","GBX2","LRPPRC","MSH6","KCNQ2","HLCS",
           "LINC01424","BCL2L13","OGFRP1","LINC00636","VILL","CDC25A","RAD18",
           "LINC02374","ZFP42","CFAP299","NR2F1-AS1","FAM83A-AS1","FAM83A-AS1",
           "KBTBD11-OT1","DNAJC5B","MIR8055","LOC101926892","RALGPS1","UNQ6494"),
  delta_beta = c(0.1786673,0.0658881,0.1003922,0.0927014,0.0930926,0.1447595,
                 0.1581512,0.1673264,0.1464119,0.1662389,0.0943406,0.1426029,
                 0.1347748,-0.1132179,0.1220984,0.1504201,0.1931537,0.1787610,
                 0.1203571,0.1428903,0.1116022,-0.1575064,0.0611679,0.0904769,
                 0.1531401,0.0774243,0.1617446,0.1424714,0.0794246,0.2312240,
                 0.0582721,0.0773780,0.1531674,0.1269485,-0.1246056,0.1627953,
                 0.1263904,0.1855893,0.1350595,0.1594606,0.1168616,0.1236535,
                 0.0980484,-0.1476910,-0.0812841,0.1052451,0.0858787,0.2017944),
  PValue = c(0.0006797,0.0009258,0.0007340,0.0006388,0.0008428,0.0004245,
             0.0000681,0.0000584,0.0000987,0.0001942,0.0004118,0.0007074,
             0.0005422,0.0007757,0.0000690,0.0000226,0.0000233,0.0006619,
             0.0001442,0.0003776,0.0001143,0.0004920,0.0003446,0.0005360,
             0.0007096,0.0004372,0.0003874,0.0000767,0.0001617,0.0004739,
             0.0009016,0.0000800,0.0000918,0.0008229,0.0007412,0.0008259,
             0.0002025,0.0001020,0.0003266,0.0006677,0.0005007,0.0007297,
             0.0004791,0.0006030,0.0003167,0.0003597,0.0002375,0.0006420),
  FDR = c(0.6531798,0.6743415,0.6531798,0.6531798,0.6531798,0.6423066,
          0.5380918,0.5380918,0.5401654,0.5574120,0.6423066,0.6531798,
          0.6531798,0.6531798,0.5380918,0.3971365,0.3971365,0.6531798,
          0.5401654,0.6423066,0.5401654,0.6515627,0.6407444,0.6531798,
          0.6531798,0.6442671,0.6423066,0.5380918,0.5401654,0.6442671,
          0.6621912,0.5380918,0.5401654,0.6531798,0.6531798,0.6531798,
          0.5574120,0.5401654,0.6407444,0.6531798,0.6531798,0.6531798,
          0.6442671,0.6531798,0.6407444,0.6415295,0.5852835,0.6531798)
) %>%
  mutate(direction = ifelse(delta_beta > 0, "Hyper", "Hypo"),
         source    = "Whole-genome locus-level (exploratory)")

# ── 3. Combine, deduplicate (keep most significant record per gene) ────────
candidate_dml_genes <- bind_rows(tss_filtered, wg_filtered) %>%
  group_by(gene) %>%
  slice_min(order_by = coalesce(PValue, Inf), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Total unique candidate DML genes (TSS-filtered + whole-genome):",
    nrow(candidate_dml_genes), "\n\n")
print(candidate_dml_genes, n = Inf)

# ── 4. Cross-reference against RNA-seq DEGs ─────────────────────────────────
rnaseq_lookup <- res_df %>%
  filter(Gene.name %in% candidate_dml_genes$gene) %>%
  dplyr::select(gene = Gene.name, logFC, FDR_rna = FDR) %>%
  mutate(is_DEG = FDR_rna < 0.05 & abs(logFC) > log2(1.5))

merged <- candidate_dml_genes %>%
  left_join(rnaseq_lookup, by = "gene")

cat("\nCandidate DML genes with a matching RNA-seq record:",
    sum(!is.na(merged$logFC)), "/", nrow(merged), "\n")
cat("Of these, formal DEGs (FDR<0.05, |FC|>1.5):",
    sum(merged$is_DEG, na.rm = TRUE), "\n\n")

print(merged %>% arrange(desc(is_DEG), delta_beta), n = Inf)
```

## ── snoRNA artefact diagnostics (Tim's comment) ────────────────────────────

```{r}

## (i) Library size / composition check — RIN not present in your clinical
## metadata, so this uses library size and % of reads mapping to snoRNA
## biotypes as the available proxies for RNA degradation/capture bias.

lib_size_df <- data.frame(
  subject   = y$samples$subject,
  group     = y$samples$group,
  lib_size  = y$samples$lib.size
)

cat("Library size by group:\n")
print(tapply(lib_size_df$lib_size, lib_size_df$group, summary))

wilcox_libsize <- wilcox.test(lib_size ~ group, data = lib_size_df)
cat("\nWilcoxon test, library size DG vs NG: p =", format(wilcox_libsize$p.value, digits = 3), "\n")

ggplot(lib_size_df, aes(x = group, y = lib_size, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.7) +
  geom_jitter(aes(colour = group), width = 0.12, size = 2.5, alpha = 0.85) +
  scale_fill_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0")) +
  scale_colour_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0")) +
  labs(title = "Library size by group", x = NULL, y = "Library size (total counts)") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")

ggsave("11.06.run/libsize_by_group.pdf", width = 5, height = 5)

## % of total library mapping to snoRNA/snRNA/scaRNA biotypes per sample —
## a rise in this percentage in one group is a classic degraded-RNA /
## capture-bias signature, independent of any single gene's fold-change.

snorna_ensembl <- res_df$Row.names[grepl("^SNORA|^SNORD|^SCARNA", res_df$Gene.name)]
# adjust the ID column name above if res_df's ENSEMBL IDs live in rownames
# instead of a Row.names column — check with head(res_df) first

raw_counts <- y$counts
pct_snorna <- data.frame(
  subject    = y$samples$subject,
  group      = y$samples$group,
  pct_snorna = 100 * colSums(raw_counts[rownames(raw_counts) %in% snorna_ensembl, ]) /
                     colSums(raw_counts)
)

cat("\n% of library mapping to snoRNA/snRNA/scaRNA biotypes, by group:\n")
print(tapply(pct_snorna$pct_snorna, pct_snorna$group, summary))

wilcox_pctsnorna <- wilcox.test(pct_snorna ~ group, data = pct_snorna)
cat("\nWilcoxon test, %snoRNA reads DG vs NG: p =", format(wilcox_pctsnorna$p.value, digits = 3), "\n")

ggplot(pct_snorna, aes(x = group, y = pct_snorna, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.7) +
  geom_jitter(aes(colour = group), width = 0.12, size = 2.5, alpha = 0.85) +
  scale_fill_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0")) +
  scale_colour_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0")) +
  labs(title = "% library mapping to snoRNA/snRNA/scaRNA biotypes",
       x = NULL, y = "% of total library") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")

ggsave("11.06.run/pct_snorna_by_group.pdf", width = 5, height = 5)


## (ii) Is the signal driven by a handful of outlier samples? ───────────────
## Reuses your existing logcpm object and top DEG list, styled like your
## SNORA26 boxplot chunk further down in the notebook.

top_snorna_genes <- res_df %>%
  filter(FDR < 0.05 & logFC > log2(1.5)) %>%
  filter(grepl("^SNORA|^SNORD|^SCARNA", Gene.name)) %>%
  arrange(FDR) %>%
  slice_head(n = 6) %>%
  pull(Gene.name)

logcpm <- cpm(y, log = TRUE, prior.count = 1)

snorna_long <- res_df %>%
  filter(Gene.name %in% top_snorna_genes) %>%
  pull(Gene.name)

# match Gene.name back to rownames of logcpm via res_df
gene_to_id <- setNames(rownames(res_df), res_df$Gene.name)
snorna_ids <- gene_to_id[top_snorna_genes]

snorna_expr <- as.data.frame(t(logcpm[snorna_ids, ]))
colnames(snorna_expr) <- top_snorna_genes
snorna_expr$subject <- y$samples$subject
snorna_expr$group   <- y$samples$group

snorna_long_df <- snorna_expr %>%
  tidyr::pivot_longer(cols = all_of(top_snorna_genes),
                      names_to = "gene", values_to = "logCPM")

ggplot(snorna_long_df, aes(x = group, y = logCPM)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.5) +
  geom_jitter(aes(colour = group), width = 0.12, size = 2, alpha = 0.85) +
  facet_wrap(~gene, scales = "free_y") +
  scale_colour_manual(values = c("DG" = "#D94F3D", "NG" = "#4C72B0")) +
  labs(title = "Per-sample expression of top upregulated snoRNAs",
       x = NULL, y = "log2-CPM") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave("11.06.run/snorna_per_sample_outlier_check.pdf", width = 9, height = 6)


## (iii) Does the enrichment persist after excluding snoRNA/snRNA/scaRNA
## biotypes from the DE model entirely? ──────────────────────────────────
## Re-runs your Section 3/8 pipeline (filter → normalise → fit → LRT)
## on a biotype-excluded gene set, for direct comparison against res_df.

is_snorna_biotype <- grepl("^SNORA|^SNORD|^SCARNA", genes[rownames(y), "Gene.name"])
cat("\nGenes flagged as snoRNA/snRNA/scaRNA biotype in filtered set:",
    sum(is_snorna_biotype), "\n")

y_nosnorna <- y[!is_snorna_biotype, ]
y_nosnorna <- calcNormFactors(y_nosnorna)

design_nosnorna <- model.matrix(~0 + y_nosnorna$samples$group)
colnames(design_nosnorna) <- c("DG", "NG")

y_nosnorna   <- estimateDisp(y_nosnorna, design_nosnorna)
fit_nosnorna <- glmFit(y_nosnorna, design_nosnorna)
lrt_nosnorna <- glmLRT(fit_nosnorna, contrast = c(1, -1))

res_nosnorna <- as.data.frame(topTags(lrt_nosnorna, n = Inf)$table)

deg_nosnorna <- res_nosnorna %>% filter(FDR < 0.05 & abs(logFC) > log2(1.5))
cat("DEGs after excluding snoRNA/snRNA/scaRNA biotypes:", nrow(deg_nosnorna), "\n")
cat("  Upregulated:  ", sum(deg_nosnorna$logFC > 0), "\n")
cat("  Downregulated:", sum(deg_nosnorna$logFC < 0), "\n")

## Check whether your other headline DEGs are stable without snoRNAs
headline_genes <- c("PPARGC1A", "FGFR2", "ADIPOQ", "INSR", "IRS1")
cat("\nHeadline gene stability check (with vs without snoRNA biotypes):\n")
for (g in headline_genes) {
  with_row    <- res_df[res_df$Gene.name %in% g, c("logFC", "FDR")]
  without_row <- res_nosnorna[res_nosnorna$Gene.name %in% g, c("logFC", "FDR")]

  # Guard in case a gene genuinely has zero or >1 matches
  if (nrow(with_row) != 1 || nrow(without_row) != 1) {
    cat(sprintf("  %-10s  WARNING: %d match(es) in res_df, %d match(es) in res_nosnorna — skipping\n",
                g, nrow(with_row), nrow(without_row)))
    next
  }

  cat(sprintf("  %-10s  with: logFC=%.3f FDR=%.2e  |  without: logFC=%.3f FDR=%.2e\n",
              g, with_row$logFC, with_row$FDR, without_row$logFC, without_row$FDR))
}
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# ── Use corrected threshold ────────────────────────────────────────────────
sig_up   <- res_df[res_df$FDR < 0.05 & res_df$logFC > 0, ]
sig_down <- res_df[res_df$FDR < 0.05 & res_df$logFC < 0, ]

cat("Upregulated DEGs:  ", nrow(sig_up), "\n")
cat("Downregulated DEGs:", nrow(sig_down), "\n")

# ── Map ENSEMBL to Entrez ─────────────────────────────────────────────────
map_entrez <- function(df) {
  df$ENSEMBL <- rownames(df)
  entrez <- bitr(df$ENSEMBL, fromType = "ENSEMBL",
                 toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  merge(df, entrez, by = "ENSEMBL")
}

up_mapped   <- map_entrez(sig_up)
down_mapped <- map_entrez(sig_down)

# ── KEGG enrichment ───────────────────────────────────────────────────────
kegg_up <- enrichKEGG(gene = up_mapped$ENTREZID, organism = "hsa",
                      pvalueCutoff = 0.05, qvalueCutoff = 0.2)

kegg_down <- enrichKEGG(gene = down_mapped$ENTREZID, organism = "hsa",
                        pvalueCutoff = 0.05, qvalueCutoff = 0.2)

# ── GO:BP enrichment ──────────────────────────────────────────────────────
go_up <- enrichGO(gene = up_mapped$ENTREZID, OrgDb = org.Hs.eg.db,
                  ont = "BP", pvalueCutoff = 0.05, qvalueCutoff = 0.2,
                  readable = TRUE)

go_down <- enrichGO(gene = down_mapped$ENTREZID, OrgDb = org.Hs.eg.db,
                    ont = "BP", pvalueCutoff = 0.05, qvalueCutoff = 0.2,
                    readable = TRUE)

# ── Semantic redundancy reduction for GO:BP ────────────────────────────────
# GO's ontology structure means parent/child/sibling terms describing the
# same underlying biology (e.g. "cellular respiration" and "aerobic
# respiration") frequently all pass significance together. simplify()
# collapses these to representative terms using semantic similarity
# (cutoff = 0.7 is clusterProfiler's standard default), keeping the term
# with the lowest p.adjust from each redundant cluster. This is applied
# to GO only — KEGG has no equivalent hierarchy/similarity structure to
# collapse on, so its redundancy (see note below) is handled separately.
go_up_simplified   <- clusterProfiler::simplify(go_up,   cutoff = 0.7, by = "p.adjust", select_fun = min)
go_down_simplified <- clusterProfiler::simplify(go_down, cutoff = 0.7, by = "p.adjust", select_fun = min)

cat("\nGO:BP Upregulated — before simplify():", nrow(as.data.frame(go_up)),
    "| after:", nrow(as.data.frame(go_up_simplified)), "terms\n")
cat("GO:BP Downregulated — before simplify():", nrow(as.data.frame(go_down)),
    "| after:", nrow(as.data.frame(go_down_simplified)), "terms\n")

# ── KEGG redundancy flag: MHC class II gene module ─────────────────────────
# Several KEGG "disease" pathways (Systemic lupus erythematosus, Type I
# diabetes mellitus, Leishmaniasis, Toxoplasmosis, Viral myocarditis, etc.)
# are populated almost entirely by the same HLA-DR/DP/DQ gene set, since
# KEGG catalogues antigen-presentation genes under disease-pathway names
# rather than a single generic "MHC class II presentation" term. These are
# NOT five independent biological signals — flag rather than silently plot
# all of them as if they were.
mhc_genes <- c("HLA-DPA1", "HLA-DPB1", "HLA-DQB1", "HLA-DRA", "HLA-DRB1", "HLA-DRB5")

kegg_up_df <- as.data.frame(kegg_up)
kegg_up_df$mhc_driven <- sapply(strsplit(kegg_up_df$geneID, "/"), function(entrez_ids) {
  syms <- mapIds(org.Hs.eg.db, keys = entrez_ids, column = "SYMBOL",
                 keytype = "ENTREZID", multiVals = "first")
  mean(syms %in% mhc_genes) > 0.5   # >50% of the pathway's genes are MHC-II
})

cat("\nKEGG Upregulated pathways flagged as MHC class II-driven (>50% overlap):\n")
print(kegg_up_df$Description[kegg_up_df$mhc_driven])

# ── Print summaries ───────────────────────────────────────────────────────
cat("\nKEGG - Upregulated:", nrow(as.data.frame(kegg_up)), "pathways\n")
cat("KEGG - Downregulated:", nrow(as.data.frame(kegg_down)), "pathways\n")
cat("GO:BP - Upregulated (simplified):", nrow(as.data.frame(go_up_simplified)), "terms\n")
cat("GO:BP - Downregulated (simplified):", nrow(as.data.frame(go_down_simplified)), "terms\n")

# ── Export ────────────────────────────────────────────────────────────────
write.table(kegg_up_df,                             "11.06.run/KEGG_upregulated.txt",    sep = "\t", quote = FALSE, row.names = FALSE)
write.table(as.data.frame(kegg_down),                "11.06.run/KEGG_downregulated.txt",  sep = "\t", quote = FALSE, row.names = FALSE)
write.table(as.data.frame(go_up_simplified),          "11.06.run/GO_BP_upregulated_simplified.txt",   sep = "\t", quote = FALSE, row.names = FALSE)
write.table(as.data.frame(go_down_simplified),        "11.06.run/GO_BP_downregulated_simplified.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r}
library(enrichplot)

p_kegg_up_dot <- dotplot(kegg_up, showCategory = 15, color = "p.adjust") +
  ggtitle("KEGG — Upregulated (DG vs NG)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_kegg_down_dot <- dotplot(kegg_down, showCategory = 15, color = "p.adjust") +
  ggtitle("KEGG — Downregulated (DG vs NG)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(p_kegg_up_dot)
print(p_kegg_down_dot)

ggsave("11.06.run/KEGG_dotplot_up.pdf", p_kegg_up_dot, width = 8, height = 7)
ggsave("11.06.run/KEGG_dotplot_down.pdf", p_kegg_down_dot, width = 8, height = 7)
```

```{r}
library(ggplot2)
library(dplyr)
library(stringr)

kegg_combined <- bind_rows(
  as.data.frame(kegg_up)   %>% mutate(direction = "Upregulated"),
  as.data.frame(kegg_down) %>% mutate(direction = "Downregulated")
) %>%
  group_by(direction) %>%
  arrange(p.adjust) %>%
  slice_head(n = 15) %>%
  ungroup() %>%
  mutate(
    Description = str_wrap(Description, width = 45),
    gene_ratio  = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1]) / as.numeric(x[2])),
    gene_count  = as.numeric(sub("/.*", "", GeneRatio)),
    Description = factor(Description, levels = rev(unique(Description[order(direction, p.adjust)])))
  )

p_kegg_dot <- ggplot(kegg_combined, aes(x = gene_ratio, y = Description)) +
  geom_point(aes(size = gene_count, colour = direction), alpha = 0.85) +
  facet_wrap(~direction, scales = "free_y") +
  scale_colour_manual(values = c("Upregulated" = DG_COL, "Downregulated" = NG_COL), name = NULL) +
  scale_size_continuous(name = "Gene count", range = c(3, 9)) +
  labs(title = "KEGG pathway enrichment — DG vs NG",
       x = "Gene ratio", y = NULL) +
  theme_classic(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "#2C3E50", colour = NA),
    strip.text        = element_text(colour = "white", face = "bold"),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.y          = element_text(size = 9),
    legend.position        = "right"
  )

print(p_kegg_dot)
ggsave("11.06.run/KEGG_dotplot_combined.pdf", p_kegg_dot, width = 12, height = 7)
```
